Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit

### 2024-06-19 optim.qmd
more tweaks

@tbl-smoothness-squintability presents the parameters estimated from the
Gaussian process (variance, range, smooth, and nugget) and the scaled
logistic function ($\theta_1$ to $\theta_4$) for each case. The column
"smooth" is used as the smoothness measure and the column "squint" is
calculated as [TODO: equation reference] as squintability.
The low squint value of MIC and TIC index is due to its convex shape of
the index value against the projection distance, as shown in
@fig-idx-proj-dist. Comparing to other concave shapes, the maximum first
derivative happens at a smaller projection distance ($\theta_2$),
requiring the optimiser to get closer to the optimal to see a
significant change in the index value, hence a small squintability
measure.

```{r fig-idx-proj-dist}
#| fig.width = 9,
#| fig.height = 5,
#| fig.cap = "Index values versus projection distance for the 12 pipe/sine-wave finding problem, after the binning procedure for calculating the squintability measure. The index values, averaged at bin width of 0.005, are scaled from 0 to 1 for comparison (holes, TIC, and stringy). The `MIC` and `TIC` index curves are convex while others are concave. The stringy curve shows an instantaneous jump to the optimum when aproaching the best basis." 
sq_basis_dist_idx |>
  ggplot(aes(x = dist, y = idx, group = index)) +
  geom_line() +
  facet_nested_wrap(~index + n, nrow = 3, labeller = label_both) + 
  xlab("projection distance") + ylab("index value")

@@

```{r}
#| label: tbl-mod-output
#| tbl-cap: Model estimates of proportion of jellyfish success on optimisation properties and jellyfish hyper-parameters from the generalised linear model with a binomial family and a logit link function. The variable smoothness, squintability, number of jellyfish and maximum number of tries are positively associated with the JS optimiser success rate while data dimension and being flagged as long runtime are negatively associated with the success rate. The variable squintability and dimension are significant, suggesting their importance relative to jellyfish hyper-parameters in the optimisation success.
sim_df2 <- sim_df |>  
  mutate(n_jellies = n_jellies/10,
         max_tries = max_tries/10,
         long_time = ifelse(time > 30, 1, 0),
         P_J = ifelse(index == "stringy", 0, P_J))

>>>>

@tbl-smoothness-squintability presents the parameters estimated from the
Gaussian process (variance, range, smooth, and nugget) and the scaled
logistic function ($\theta_1$ to $\theta_4$) for each case. The column
"smooth" is used as the smoothness measure and the column "squint" is
calculated as \[TODO: equation reference\] as squintability. The low
squint value of MIC and TIC index is due to its convex shape of the
index value against the projection distance, as shown in
@fig-idx-proj-dist. Comparing to other concave shapes, the maximum first
derivative happens at a smaller projection distance ($\theta_2$),
requiring the optimiser to get closer to the optimal to see a
significant change in the index value, hence a small squintability
measure.

```{r fig-idx-proj-dist}
#| fig.width = 9,
#| fig.height = 5,
#| fig.cap = "Index values versus projection distance for the 12 pipe/sine-wave finding problem, after the binning procedure for calculating the squintability measure. The index values, averaged at bin width of 0.005, are scaled from 0 to 1 for comparison (holes, TIC, and stringy). The `MIC` and `TIC` index curves are convex while others are concave. The stringy curve shows an instantaneous jump to the optimum when aproaching the best basis." 
sq_basis_dist_idx |>
  ggplot(aes(x = dist, y = index, group = index_name)) +
  geom_point() +
  facet_nested_wrap(~index + n, nrow = 3, labeller = label_both) + 
  xlab("projection distance") + ylab("index value") 
```

@@

```{r}
#| label: tbl-mod-output
#| tbl-cap: Model estimates of proportion of jellyfish success on optimisation properties and jellyfish hyper-parameters from the generalised linear model with a binomial family and a logit link function. The variable smoothness, squintability, number of jellyfish and maximum number of tries are positively associated with the JS optimiser success rate while data dimension and being flagged as long runtime are negatively associated with the success rate. The variable squintability and dimension are significant, suggesting their importance relative to jellyfish hyper-parameters in the optimisation success.
sim_df2 <- sim_df |>   
  mutate(n_jellies = n_jellies/10,
         max_tries = max_tries/10,
         long_time = ifelse(time > 20, 1, 0),
         P_J = ifelse(index == "stringy", 0, P_J))

### 2024-06-19 optim.tex
more tweaks

\begin{tcolorbox}[enhanced jigsaw, bottomrule=.15mm, toptitle=1mm, left=2mm, opacitybacktitle=0.6, colbacktitle=quarto-callout-note-color!10!white, bottomtitle=1mm, titlerule=0mm, toprule=.15mm, opacityback=0, colframe=quarto-callout-note-color-frame, title={Algorithm: Jellyfish Optimizer Pseudo Code}, colback=white, breakable, arc=.35mm, rightrule=.15mm, leftrule=.75mm, coltitle=black]

@@

In addition to the pipe-finding problem, a new shape, sine wave, is
investigated in 6D and 8D spaces with six indices considered:
\texttt{dcor2d\_2}, \texttt{loess2d}, \texttt{MIC}, \texttt{TIC},
\texttt{spline}, and \texttt{stringy}. Combining with two jellyfish
hyper-parameters, a total of 42 cases is produced, comprising of 24
pipe-finding cases and 8 sine-wave finding cases. For each case, the JS
optimiser is run fifty times and the success rate is calculated as in
Section~\ref{sec-app-1}.

@@

\begin{tabular}{|>{}lll>{}r|rrr>{}r|rrr>{}r|>{}r|}
\toprule
  & shape & index & d & variance & range & smooth & nugget & $\theta_1$ & $\theta_2$ & $\theta_3$ & $\theta_4$ & squint\\
\midrule
1 & pipe & holes & 6 & 0.002 & 0.408 & 2.364 & 0.212 & 0.991 & 0.860 & 3.368 & 0.172 & 0.627\\
2 & pipe & holes & 8 & 0.000 & 0.259 & 2.373 & 0.613 & 1.011 & 0.869 & 3.264 & 0.085 & 0.696\\
3 & pipe & holes & 10 & 0.000 & 0.144 & 2.317 & 1.831 & 1.011 & 0.885 & 3.151 & 0.153 & 0.636\\
4 & pipe & holes & 12 & 0.000 & 0.254 & 2.173 & 0.879 & 0.999 & 0.878 & 3.345 & 0.196 & 0.621\\
5 & sine & MIC & 6 & 0.016 & 0.100 & 2.394 & 0.087 & 0.894 & 0.571 & 1.623 & -0.024 & 0.314\\
\addlinespace
6 & sine & MIC & 8 & 0.016 & 0.100 & 2.394 & 0.087 & 0.932 & 0.328 & 1.314 & -0.030 & 0.193\\
7 & sine & TIC & 6 & 0.124 & 0.104 & 2.471 & 0.086 & 0.951 & 0.536 & 1.719 & -0.025 & 0.330\\
8 & sine & TIC & 8 & 0.124 & 0.104 & 2.471 & 0.086 & 0.945 & 0.564 & 1.723 & -0.027 & 0.341\\
9 & sine & dcor2d & 6 & 0.034 & 0.167 & 2.663 & 0.114 & 0.954 & 1.039 & 2.742 & -0.019 & 0.737\\
10 & sine & loess2d & 6 & 0.083 & 0.307 & 2.194 & 0.292 & 1.016 & 1.039 & 2.648 & 0.080 & 0.689\\
\addlinespace
11 & sine & splines2d & 6 & 0.040 & 0.189 & 2.606 & 0.104 & 1.014 & 1.051 & 2.730 & -0.009 & 0.780\\
12 & sine & stringy & 6 & 0.000 & 1173.035 & 1.031 & 17608.047 & 1.040 & 0.011 & 254.733 & 0.170 & 0.648\\
\bottomrule
\end{tabular}
\endgroup{}

@@

\toprule
Index & D & Smoothness & Squintability & Num. of jellyfish & Max. Num. of tries & Prop. of success & Time\\
\midrule
dcor2d\_2 & 6 & 2.663 & 0.737 & 20 & 50 & 0.44 & 35.320 secs\\
dcor2d\_2 & 6 & 2.663 & 0.737 & 20 & 100 & 0.60 & 70.347 secs\\
dcor2d\_2 & 6 & 2.663 & 0.737 & 50 & 50 & 0.90 & 87.158 secs\\
holes & 6 & 2.364 & 0.627 & 20 & 50 & 0.88 & 0.801 secs\\
holes & 6 & 2.364 & 0.627 & 20 & 100 & 0.96 & 1.456 secs\\
\addlinespace
holes & 6 & 2.364 & 0.627 & 50 & 50 & 1.00 & 1.477 secs\\
holes & 6 & 2.364 & 0.627 & 50 & 100 & 1.00 & 3.052 secs\\
\bottomrule

@@

\toprule
term & estimate & std.error & statistic & p.value\\
\midrule
Intercept & 2.685 & 15.464 & 0.174 & 0.862\\
Smoothness & 3.070 & 3.631 & 0.845 & 0.398\\
Squintability & -8.545 & 22.252 & -0.384 & 0.701\\
dimension (d) & -0.610 & 0.355 & -1.719 & 0.086\\
long time & -1.683 & 1.930 & -0.872 & 0.383\\
\addlinespace
number of jellyfish & 0.269 & 0.175 & 1.538 & 0.124\\
maximum number of tries & 0.080 & 0.196 & 0.410 & 0.682\\
\bottomrule

>>>>

\begin{tcolorbox}[enhanced jigsaw, title={Algorithm: Jellyfish Optimizer Pseudo Code}, arc=.35mm, left=2mm, breakable, toprule=.15mm, colframe=quarto-callout-note-color-frame, opacityback=0, titlerule=0mm, bottomtitle=1mm, colback=white, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, leftrule=.75mm, colbacktitle=quarto-callout-note-color!10!white, opacitybacktitle=0.6, coltitle=black]

@@

In addition to the pipe-finding problem, a new shape, sine wave, is
investigated in 6D and 8D spaces with six indices considered:
\texttt{dcor2d\_2}, \texttt{loess2d}, \texttt{MIC}, \texttt{TIC},
\texttt{spline}, and \texttt{stringy}. Combining with two jellyfish
hyper-parameters, a total of 52 cases is produced, comprising of 24
pipe-finding cases and 28 sine-wave finding cases. For each case, the JS
optimiser is run fifty times and the success rate is calculated as in
Section~\ref{sec-app-1}.

@@

\begin{tabular}{|>{}lll>{}r|rrr>{}r|rrr>{}r|>{}r|}
\toprule
  & shape & index & d & variance & range & smooth & nugget & $\theta_1$ & $\theta_2$ & $\theta_3$ & $\theta_4$ & squint\\
\midrule
1 & pipe & holes & 6 & 0.002 & 0.371 & 2.355 & 0.221 & 1.015 & 0.860 & 3.368 & 0.018 & 0.763\\
2 & pipe & holes & 8 & 0.000 & 0.178 & 2.189 & 0.820 & 1.014 & 0.869 & 3.264 & 0.029 & 0.740\\
3 & pipe & holes & 10 & 0.000 & 0.107 & 2.192 & 3.027 & 1.016 & 0.885 & 3.151 & 0.022 & 0.737\\
4 & pipe & holes & 12 & 0.000 & 0.148 & 2.295 & 1.581 & 1.011 & 0.878 & 3.345 & 0.004 & 0.779\\
5 & sine & MIC & 6 & 0.016 & 0.092 & 2.457 & 0.083 & 0.894 & 0.571 & 1.623 & -0.024 & 0.314\\
\addlinespace
6 & sine & MIC & 8 & 0.016 & 0.081 & 2.636 & 0.084 & 0.932 & 0.328 & 1.314 & -0.030 & 0.193\\
7 & sine & TIC & 6 & 0.124 & 0.110 & 2.444 & 0.086 & 0.951 & 0.536 & 1.719 & -0.027 & 0.330\\
8 & sine & TIC & 8 & 0.125 & 0.099 & 2.475 & 0.085 & 0.949 & 0.564 & 1.723 & -0.028 & 0.342\\
9 & sine & dcor2d & 6 & 0.034 & 0.167 & 2.663 & 0.114 & 0.954 & 1.039 & 2.742 & -0.019 & 0.737\\
10 & sine & loess2d & 6 & 0.079 & 0.336 & 1.988 & 0.309 & 1.016 & 1.039 & 2.648 & 0.080 & 0.689\\
\addlinespace
11 & sine & splines2d & 6 & 0.042 & 0.242 & 2.537 & 0.102 & 1.014 & 1.051 & 2.730 & -0.009 & 0.780\\
12 & sine & stringy & 6 & 0.000 & 0.007 & 1.536 & 38.166 & 1.045 & 0.011 & 254.734 & 0.053 & 0.739\\
\bottomrule
\end{tabular}
\endgroup{}

@@

\toprule
Index & D & Smoothness & Squintability & Num. of jellyfish & Max. Num. of tries & Prop. of success & Time\\
\midrule
MIC & 6 & 2.457 & 0.314 & 20 & 50 & 0.12 & 2.479 secs\\
MIC & 6 & 2.457 & 0.314 & 20 & 100 & 0.24 & 8.950 secs\\
MIC & 6 & 2.457 & 0.314 & 50 & 50 & 0.52 & 5.651 secs\\
MIC & 6 & 2.457 & 0.314 & 50 & 100 & 0.64 & 13.223 secs\\
MIC & 6 & 2.457 & 0.314 & 100 & 50 & 0.76 & 19.453 secs\\
\addlinespace
MIC & 8 & 2.636 & 0.193 & 20 & 50 & 0.08 & 2.566 secs\\
MIC & 8 & 2.636 & 0.193 & 20 & 100 & 0.08 & 4.960 secs\\
\bottomrule

@@

\toprule
term & estimate & std.error & statistic & p.value\\
\midrule
Intercept & -4.523 & 5.332 & -0.848 & 0.396\\
Smoothness & 1.195 & 1.915 & 0.624 & 0.533\\
Squintability & 8.241 & 2.737 & 3.011 & 0.003\\
dimension (d) & -0.628 & 0.255 & -2.462 & 0.014\\
long time & -0.874 & 1.286 & -0.680 & 0.497\\
\addlinespace
number of jellyfish & 0.216 & 0.127 & 1.701 & 0.089\\
maximum number of tries & 0.113 & 0.150 & 0.752 & 0.452\\
\bottomrule
